Data cleaning and processing script

### 2023-03-14 code/analysis_code/exploratory/exploratory_analysis.qmd
general cleaning

---
title: "Exploratory analysis of year 2 surveillance data"
author: "Leah Lariscy"
format: html
editor: visual
theme: solar
---

@@

```{r}
here()
n1_n2_clean <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
plant_clean <- readRDS(here("data/processed_data/year2_plant_clean.rds"))
n1_n2_plant_combine <- readRDS(here("data/processed_data/year2_cfx_plant_combined.rds"))
covid_combine <- readRDS(here("data/processed_data/all_covid_combined.rds"))
all_data_combine <- readRDS(here("data/processed_data/year2_all_data_combined.rds"))
recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))
n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))

@@

### Case 7 day moving average overtime, bar plot and GAM model

@@

### PCR positives per day over time, bar plot and GAM model

@@

   ggplot(aes(x = date, y = mean_14_dma)) +
  geom_bar(stat = "identity", color = '#FF7E5A') +
  stat_smooth(method = "gam", color = 'black') +
  ylab(label = "Clinical Case 14-Day Moving Average") +
  xlab(label = "Date") +
  theme_light()

@@

### Log10 total copies in wastewater over time (avg N1/N2)

@@

### Log10 viral load in wastewater for N1 and N2

@@

@@

### Log10 copies per liter vs case 7DMA (N1/N2 average)

@@

### Log10 copies per liter vs case 7DMA (N1 & N2)

@@

### Log10 Copies per Liter vs daily clinical cases

@@

### Log10 Copies per Liter vs daily clinical cases (N1 N2 average)

@@

### Log10 viral load vs 7DMA

@@

### Log10 viral load vs 7DMA (N1 & N2)

@@

### Look at relationship of copies/L and 7dma over time

```{r}
all_data_combine %>% #filter(target == "N2") %>% 
  drop_na() %>% 
  group_by(date, facility) %>% 
  summarize(mean_copies_L = mean(avg_copy_L), 
            mean_7dma = mean(pcr_pos_7dma)) %>% 
  ungroup(facility) %>% 
  summarize(sum_mean_copies_L = sum(mean_copies_L),
            mean_7dma = mean(mean_7dma)) %>% 
  ggplot(aes(x=date)) +
  #geom_point(aes(y=log10(sum_mean_copies_L)/mean_7dma)) + 
  geom_line(aes(y=(log10(sum_mean_copies_L)/mean_7dma), color = "red")) +
  geom_line(aes(y=log10(mean_7dma)), color = "blue")

### Log10 viral load vs daily positive cases

@@

### Log10 viral load vs daily positive cases

@@

>>>>

---
title: "Exploratory analysis of year 2 surveillance data"
author: "Leah Lariscy"
editor: visual
format:
  html:
    toc: true
    toc-depth: 3
theme: solar
---

@@

```{r}
here()
n1_n2_clean <- readRDS(here("data/processed_data/year2_cfx_clean.rds"))
plant_clean <- readRDS(here("data/processed_data/year2_plant_clean.rds"))
n1_n2_plant_combine <- readRDS(here("data/processed_data/year2_cfx_plant_combined.rds"))
covid_combine <- readRDS(here("data/processed_data/all_covid_combined.rds"))
all_data_combine <- readRDS(here("data/processed_data/year2_all_data_combined.rds"))
recovery_percent <- readRDS(here("data/processed_data/recovery_percentages.rds"))
n1_n2n2_avg <- readRDS(here("data/processed_data/year2_cfx_avg.rds"))
prop_pos_bio <- readRDS(here("data/processed_data/prop_pos_bio.rds"))
prop_pos_sample <- readRDS(here("data/processed_data/prop_pos_sample.rds"))

@@

### Case 7 day moving average overtime

@@

### PCR positives per day over time

@@

   ggplot(aes(x = date, y = mean_14_dma)) +
  geom_bar(stat = "identity", color = '#FF7E5A') +
  geom_line() +
  ylab(label = "Clinical Case 14-Day Moving Average") +
  xlab(label = "Date") +
  theme_light()

@@

### Log10 viral load in wastewater over time (avg N1/N2)

@@

### Log10 viral load in wastewater (both N1 and N2)

@@

### Percent positive technical replicates per sample over time

```{r}
prop_pos_sample %>% group_by(date) %>% 
  summarize(avg_pos_rate = mean(pos_rate_tr)) %>%

  ggplot(aes(date, avg_pos_rate)) +
  geom_point() +
  geom_line()

@@

### Log10 copies per liter vs log10 case 7DMA (N1/N2 average)

@@

### Log10 copies per liter vs log10 case 7DMA (N1 & N2)

@@

### Log10 Copies per Liter vs log10 daily clinical cases

@@

### Log10 Copies per Liter vs log10 daily clinical cases (N1 N2 average)

@@

### Log10 viral load vs log10 7DMA

@@

### Log10 viral load vs log10 7DMA (N1 & N2)

@@

### Log10 viral load vs log10 daily positive cases

@@

### Log10 viral load vs log10 daily positive cases

@@

### Sample percent positive vs log10 7DMA

```{r}

### 2023-03-14 code/processing_code/processing.qmd
general cleaning

---
title: "Data cleaning and processing script"
author: "Leah Lariscy"
format: html
editor: visual
theme: solar
---

@@

@@

saveRDS(wbe_year2, here("data/processed_data/year2_cfx_clean.rds"))
```

@@

```{r}
wbe_year2_avg <- wbe_year2 %>% group_by(date,facility,target,collection_num) %>% 
  summarize(avg_copy_L = mean(copy_num_L), 
            avg_copy_uL_rxn = mean(copy_num_uL_rxn)) %>% 
  ungroup()
saveRDS(wbe_year2_avg, here("data/processed_data/year2_cfx_avg.rds"))
```

@@

saveRDS(plant_clean, here("data/processed_data/year2_plant_clean.rds"))
```

@@

saveRDS(n1_n2_plant, here("data/processed_data/year2_cfx_plant_combined.rds"))
```

@@

saveRDS(covid, here("data/processed_data/all_covid_combined.rds"))
```

@@

saveRDS(all_data, here("data/processed_data/year2_all_data_combined.rds"))
```

@@

```{r}
wbe_tr <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
  dplyr::summarise(
    n=n(), 
    n.miss=sum(is.na(ct)), 
    ct.mean=mean(ct,na.rm=T), 
    ct.sd=sd(ct,na.rm=T)
  ) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(n.pos = n - n.miss) %>% 
  mutate(pos = ifelse(n.pos > 0, "pos", "neg"))
wbe_tr
saveRDS(wbe_tr,here("data/processed_data/wbe_technical_reps.rds"))
```

@@

```{r}
wbe_br <- wbe_tr %>%
  group_by(date, facility, target) %>% 
  dplyr::summarise(
    n.bio = n(),
    n.bio.non.miss = sum(!is.na(ct.mean)),
    n.bio.miss = sum(is.na(ct.mean)),
    n.total = sum(n), 
    n.total.miss = sum(n.miss), 
    bio.ct.mean = mean(ct.mean, na.rm = T), 
    bio.ct.sd = sd(ct.mean, na.rm=T), 
    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
  ) %>%
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup()
wbe_br
```

@@

```{r}
wbe_samp <- wbe_br %>%
  group_by(date, facility, target) %>% 
  dplyr::summarise(
    n.pos = sum(!is.na(bio.ct.mean)),
    n.miss = sum(is.na(bio.ct.mean))) %>%
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(prop.pos = n.pos/(n.pos+n.miss)*100)
wbe_samp
```

```{r}
ggstatsplot::ggbarstats(
  data = wbe_samp,
  x = n.pos,
  y = target) +
  labs(caption = NULL)
```

>>>>

---
title: "Data cleaning and processing script"
author: "Leah Lariscy"
editor: visual
format:
  html:
    toc: true
    toc-depth: 3
theme: solar
---

@@

`wbe_year2` includes: date, facility, target, bio rep, collection number, ct, copies/uL/rxn, copies/L

@@

saveRDS(wbe_year2, here("data/processed_data/year2_cfx_clean.rds"))
wbe_year2
```

@@

`wbe_year2_avg` - copies/L is averaged across biological replicates

```{r}
wbe_year2_avg <- wbe_year2 %>% group_by(date,facility,target,collection_num) %>% 
  summarize(avg_copy_L = mean(copy_num_L), 
            avg_copy_uL_rxn = mean(copy_num_uL_rxn)) %>% 
  ungroup()
saveRDS(wbe_year2_avg, here("data/processed_data/year2_cfx_avg.rds"))
wbe_year2_avg
```

@@

saveRDS(plant_clean, here("data/processed_data/year2_plant_clean.rds"))
plant_clean
```

@@

saveRDS(n1_n2_plant, here("data/processed_data/year2_cfx_plant_combined.rds"))
n1_n2_plant
```

@@

saveRDS(covid, here("data/processed_data/all_covid_combined.rds"))
covid
```

@@

saveRDS(all_data, here("data/processed_data/year2_all_data_combined.rds"))
all_data
```

@@

This isn't super important, but I'm including anyways.

This will calculate the percent of positive technical replicates per biological replicate

```{r}
wbe_tr <- wbe_year2 %>% group_by(date, facility, target, biological_replicate) %>% 
  dplyr::summarise(
    n=n(), 
    n.miss=sum(is.na(ct)), 
    ct.mean=mean(ct,na.rm=T), 
    ct.sd=sd(ct,na.rm=T)
  ) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(n.pos = n - n.miss) %>% 
  mutate(pos = ifelse(n.pos > 0, "pos", "neg")) %>% 
  mutate(percent_pos = (n.pos/n)*100)
wbe_tr
```

@@

`pos_rate_tr` will calculate the percento of positive technical replicates, averaged across biological replicates, within each sample's targets (N1 and N2)

`percent_pos_bio` will calculate percent of positive biological replicates for each target

```{r}
wbe_br <- wbe_tr %>%
  group_by(date, facility, target) %>% 
  dplyr::summarise(
    n.bio = n(),
    n.bio.non.miss = sum(!is.na(ct.mean)),
    n.bio.miss = sum(is.na(ct.mean)),
    n.total = sum(n), 
    n.total.miss = sum(n.miss), 
    bio.ct.mean = mean(ct.mean, na.rm = T), 
    bio.ct.sd = sd(ct.mean, na.rm=T), 
    pos_rate_tr = mean(percent_pos),
    tech.ct.dists = paste(paste0(biological_replicate, " = ", round(ct.mean,2), " (sd=", round(ct.sd,2), ", n=", n-n.miss, ")"), collapse = "; ")
  ) %>%
  mutate(percent_pos_bio = (n.bio.non.miss/n.bio)*100) %>% 
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup()
wbe_br
saveRDS(wbe_br, here("data/processed_data/prop_pos_bio.rds"))
```

@@

`pos_rate_tr` calculates the percent of positive technical replicates, averaged across all biological replicates and averaged across both targets

`prop.pos` calculates percent of positive targets for each sample

```{r}
wbe_samp <- wbe_br %>%
  group_by(date, facility) %>% 
  dplyr::summarise(
    n.pos = sum(!is.na(bio.ct.mean)),
    n.miss = sum(is.na(bio.ct.mean)),
    pos_rate_tr = mean(pos_rate_tr)) %>%
  mutate_all(function(x){ifelse(is.nan(x), NA, x)}) %>% 
  ungroup() %>% 
  mutate(prop.pos = n.pos/(n.pos+n.miss)*100)
wbe_samp
saveRDS(wbe_samp, here("data/processed_data/prop_pos_sample.rds"))
```
